# Detecting Anomalies in Practice

## Operational Considerations

The World Bank Group (WBG) curates databases spanning tens of thousands
of indicators, produced by Bank units or sourced from a large number of
external providers, and covering countries and territories worldwide.
This translates to hundreds of thousands of individual time series, many
of which are updated on a regular basis. At this scale, manual review is
not feasible: anomaly detection must be as automated and systematic as
possible, which is the primary motivation for providing an open-source,
scriptable package that can be integrated into existing data pipelines.

**Using multiple methods**

Each detection method implemented in macroanomaly has its own strengths
and blind spots --- no single method reliably identifies all anomaly
types across all series. In practice, it is strongly recommended to
apply several methods in combination. Each method produces a binary flag
(anomaly or not), and some additionally provide a continuous measure of
anomalousness, such as an absolute Z-score. Running multiple methods in
parallel allows detections to be triangulated: an observation flagged by
several methods independently warrants more attention than one flagged
by only a single method.

Prioritization is a practical necessity when detection yields a large
number of flagged observations, as it frequently will at the scale of
WBG databases. Useful criteria for prioritization include the number of
methods that flagged a given observation, the magnitude of the anomaly
score, the recency of the observation, the relative importance of the
indicator or geography in question, and the reliability of the data
source.

**Feedback loops and upstream correction**

The macroanomaly package is publicly available, and the methods and
workflows described in this document are intentionally open and
replicable. This reflects a broader principle: anomalies are ideally
detected and treated at the source --- by national statistical offices,
central banks, or regional and international organizations --- rather
than corrected downstream by data distributors. When the World Bank
detects a potential anomaly in a third-party dataset, the appropriate
response is to establish a feedback loop with the originating
institution rather than to apply unilateral corrections. The tools
described here are designed to support that two-way exchange.

**The iterative nature of anomaly detection**

An important characteristic of anomaly detection that users should keep
in mind is that it is inherently iterative and relative. If a detection
script is run, some values are corrected, and the same script is then
run again on the updated series, new anomalies will often surface ---
because removing or correcting extreme values changes the mean, standard
deviation, and other distributional properties of the series, shifting
what counts as \"normal.\" This is expected behavior, not a flaw, but it
means that anomaly detection should be treated as an ongoing process
rather than a one-time procedure, and that each successive round of
detection requires the same level of scrutiny as the first.

**Human oversight is essential**

macroanomaly is a decision-support tool, not a decision-making system.
Detected anomalies must be reviewed, classified, and assessed by
qualified experts --- potentially with the assistance of AI tools such
as those described in Section 8 --- before any action is taken. Under no
circumstances should flagged values be corrected automatically based on
detection outputs alone. All changes to published data must involve
human judgment and follow established quality control and governance
procedures. The role of the package is to surface candidates for review
efficiently; the responsibility for what happens next rests with the
people using it.

## Data Preparation

Before anomaly detection can be run, the input data must meet three
requirements: it must be structured in long format, it must contain no
missing values, and it must be normalized. The following subsections
explain how to prepare a dataset that satisfies these requirements.

### Formatting the Dataset

A single input dataset passed to macroanomaly may contain multiple
indicators and multiple geographies. However, all series in the dataset
must share the same periodicity --- annual, quarterly, or monthly. If
the same indicator is available at more than one frequency, the data
should be split by periodicity and anomaly detection run separately on
each subset.

**Long vs. wide format**

Time series data is commonly stored in one of two formats: wide or long.
macroanomaly requires data in long format, so datasets stored in wide
format must be reshaped before use. The package does not provide
reshaping functions; this step should be performed using base R
(reshape()) or a package such as tidyr (pivot_longer()).

In **wide format**, each row represents a single subject --- an
indicator-geography combination --- and each time period occupies a
separate column. This produces a compact table that is easy to read at a
glance but is not well-suited to programmatic processing.

  ------------------------------------------------------------------------
  Indicator        Geography   Y2020      Y2021      Y2022      Y2023
  ---------------- ----------- ---------- ---------- ---------- ----------
  Population       Belgium     11492641   11521238   11584008   11697557

  ------------------------------------------------------------------------

In **long format**, each row represents a single observation at a single
point in time for a single subject. The table is narrower but taller,
with separate columns identifying the indicator, geography, time period,
and observed value. This is the structure macroanomaly expects.

  ------------------------------------------------------------------------
  Indicator             Geography        Year             Value
  --------------------- ---------------- ---------------- ----------------
  Population            Belgium          2020             11492641

  Population            Belgium          2021             11521238

  Population            Belgium          2022             11584008

  Population            Belgium          2023             11697557
  ------------------------------------------------------------------------

Some indicators carry additional ***dimensions***. For example,
population data disaggregated by sex would include an extra column
identifying the subgroup. Each unique combination of indicator,
dimension(s), geography, and time period corresponds to a single row.

  ------------------------------------------------------------------------
  Indicator             Sex       Geography    Year       Value
  --------------------- --------- ------------ ---------- ----------------
  Population by Sex     Total     Belgium      2020       11492641

  Population by Sex     Total     Belgium      2021       11521238

  Population by Sex     Total     Belgium      2022       11584008

  Population by Sex     Total     Belgium      2023       11697557

  Population by Sex     Male      Belgium      2020       5660064

  Population by Sex     Male      Belgium      2021       5677211

  Population by Sex     Male      Belgium      2022       5708902

  Population by Sex     Male      Belgium      2023       5761410

  Population by Sex     Female    Belgium      2020       5832577

  Population by Sex     Female    Belgium      2021       5844027

  Population by Sex     Female    Belgium      2022       5875106

  Population by Sex     Female    Belgium      2023       5936147
  ------------------------------------------------------------------------

**Required columns**

macroanomaly requires the following columns to be present in the input
dataset. Column names are flexible --- the package does not impose
specific naming conventions --- but the content of each column must
conform to the descriptions below.

- **Indicator identifier**: a code or label uniquely identifying the
  indicator.

- **Dimensions** (if applicable): one column per dimension, such as sex
  or age group.

- **Geography**: a single column identifying the country, region, or
  territory. Each geography must be represented as a single value ---
  nested or hierarchical geographic structures should be flattened
  before input.

- **Time period**: the year, quarter, or month of the observation. All
  entries in a given dataset must share the same periodicity.

- **Observed value**: the numeric value of the indicator for the given
  combination of identifier, dimensions, geography, and time period.

The combination of indicator, dimensions, geography, and time period
must be unique for every row in the dataset. Duplicate combinations will
cause errors in the detection functions and should be resolved during
data preparation.

### Imputation of Missing Values

Time series of social and economic indicators frequently contain missing
values, and most anomaly detection methods cannot handle them directly.
macroanomaly therefore imputes missing values automatically as part of
its preprocessing pipeline, via the normalize() function (called with
impute = TRUE). The imputation is performed solely to enable anomaly
detection and has no status beyond that: imputed values should never be
saved to a database or used for any purpose other than running the
detection methods.

This point deserves emphasis. Imputation for anomaly detection is
intentionally simple and neutral --- designed to fill gaps in a way that
does not distort the series\' behavior, not to produce defensible
estimates of what the missing values actually were. If gap-filled series
are needed for dissemination or analysis, a dedicated imputation
approach is required: one that is methodologically appropriate for the
indicator in question, optimized for that context, and documented in a
transparent and reproducible manner.

Because macroanomaly may flag imputed values as anomalies during
detection, the package automatically excludes such flags from its
outputs. Only anomalies detected in originally observed values are
reported and require review or treatment.

Missing data require special care because preprocessing choices can
affect anomaly scores. When imputation is used to create a continuous
series for algorithmic purposes, users should assess whether the
imputation method changes the apparent severity or location of
anomalies. Flags on imputed observations, or on observations adjacent to
long gaps, should be treated as lower-confidence signals. In operational
use, review tables and visualizations should distinguish observed from
imputed values, and users should compare results before and after
imputation when missingness is substantial. Imputed values may support
screening, but they should not be treated as equivalent to observed data
for classification or correction decisions.

**Imputation methods**

The imputation step draws on functions from the imputeTS R package. The
following methods are available:

- **na_interpolation** (default): replaces missing values using
  interpolation. The default interpolation type is linear; spline and
  Stineman interpolation are also available.

- **na_kalman**: uses a Kalman filter on a state-space model to estimate
  missing values --- well-suited to series with trend or seasonal
  structure.

- **na_mean**: replaces missing values with the series mean, optionally
  computed over a local window.

- **na_locf**: carries the last observed value forward to fill
  subsequent gaps (Last Observation Carried Forward).

- **na_ma**: replaces missing values with a weighted moving average of
  neighboring observations.

- **na_seadec**: applies seasonal decomposition before interpolating
  residuals, making it appropriate for series with strong seasonal
  patterns.

- **na_random**: fills gaps with random draws from the observed
  distribution of the series.

The default method, na_interpolation with linear interpolation, is
appropriate for most development data series and requires no parameter
tuning. For series with pronounced trends or seasonality, na_kalman or
na_seadec may produce more neutral imputations. Full documentation for
each method, including parameter options, is available in the imputeTS
package documentation.

### Detrending and Deseasonalization

Most anomaly detection methods assume that a series is approximately
stationary --- that its mean and variance remain roughly constant over
time. Social and economic time series routinely violate this assumption:
they may exhibit long-term trends (gradual upward or downward drift) and
seasonal patterns (repeating cycles tied to the calendar). If these
components are not removed before detection, the result may be
consequential: flagging observations that are unusual only relative to a
global mean when they are perfectly normal for their position in a trend
or seasonal cycle (false positives), or missing genuine anomalies that
are absorbed into an expected pattern (false negatives).

**Available methods**

A range of approaches can be used to remove trend and seasonal
components, including differencing, regression-based decomposition,
moving average filters, STL (Seasonal-Trend decomposition using LOESS),
TBATS, Prophet, and state-space models such as the Kalman filter. These
methods vary in their complexity, assumptions, and suitability for
different series characteristics.

**Implementation in macroanomaly**

Detrending and deseasonalization are handled automatically within
macroanomaly\'s normalize() function, prior to any detection step. Users
do not need to preprocess series manually. The function applies STL
decomposition by default, which is well-suited to the short, potentially
irregular series typical of development data, and returns the remainder
component as input to the detection methods. The choice of decomposition
method can be adjusted through the function\'s parameters for users who
wish to experiment with alternatives.

## Implementation Examples

We provide two examples of scripts, one with annual data, the other one
with monthly data. They can be replicated (only need to change the
default folder name). Make sure the required packages are installed (if
necessary, install them). Note that the results may vary as datasets are
updated regularly.

### Example 1: Annual Series of Education Indicators

This example uses a subset of the World Development Indicators (WDI).
The dataset is filtered to education-related indicators --- those whose
identifier begins with \"SE.\" --- and restricted to observations from
1980 onwards. As of 18 April 2026, this selection comprises 1,015,910
observations across 147 indicators and 256 geographies.

```{r}
#| eval: false

#----------------------------------------------------------------------
# Anomaly detection in WDI dataset - Standard Education (SE) indicators
#----------------------------------------------------------------------

# Load the required R packages
library(macroanomaly)
library(collapse)
library(tidyr)
library(dplyr)

# Set the default folder >>> (enter your own) <<<
setwd("C:/.../")

# Download the WDI dataset if not previously done, read, and filter (267 Mb)
wdi_url <- "https://databank.worldbank.org/data/download/WDI_CSV.zip"
if (!file.exists("WDICSV.CSV")) {
  download.file(wdi_url, destfile = "WDI.zip")
  unzip("WDI.zip")
}

df_wdi_SE <- read.csv("WDICSV.csv")
df_wdi_SE <- df_wdi_SE[startsWith(df_wdi_SE$Indicator.Code, "SE."), ]

# Reshape the dataset from wide to long
df_wdi_SE <- df_wdi_SE %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "Year", values_to = "Value") %>%
  mutate(Year = as.integer(sub("X", "", Year)))

# Drop years < 1980 and series with less than 10 non-missing values
df_wdi_SE <- df_wdi_SE[df_wdi_SE$Year >= 1980, ]
min_obs <- 10
df_wdi_SE <- df_wdi_SE %>%
  group_by(Country.Code, Indicator.Code) %>%
  filter(sum(!is.na(Value)) >= min_obs) %>%
  ungroup()

# Normalize the dataset (detrend, impute missing values)
df_wdi_SE |>
  macroanomaly::normalize(
    .value_col = "Value",
    .country_col = c("Country.Code", "Country.Name"),
    .indicator_col = "Indicator.Code",
    .time_col = "Year",
    .detrend = TRUE,
    .impute = TRUE) -> wdi_SE_normalized

# Detect anomalies (6 methods)
wdi_SE_normalized |>
  macroanomaly::detect(
    .method = c("tsoutlier", "isotree", "capa", "outliertree", "zscore", "hampel"),
    .args = list(
      capa = c(.min_seg_len = 3),
      isotree = c(threshold = 0.7),
      hampel = c(.trim = TRUE, .trim_sd = 1.5)),
    .additional_cols = TRUE) -> wdi_SE_anomalies

# Sort the anomalies data frame by severity of anomalies
wdi_SE_anomalies <- wdi_SE_anomalies[
  order(-wdi_SE_anomalies$outlier_indicator_total,
        -wdi_SE_anomalies$absZscore_zscore), ]

# Number of anomalies detected
summary(wdi_SE_anomalies)
table(wdi_SE_anomalies$outlier_indicator_total)

# Save output to CSV (all observations, with anomaly status by method)
fn <- paste0("WDI_SE_ANOMALIES", "_", format(Sys.Date(), "%Y-%m-%d"), ".CSV")
write.csv(wdi_SE_anomalies, file = fn, row.names = FALSE)

# Plot a sample of anomalies (top 20)
for (i in 1:20) {
  if (wdi_SE_anomalies$Imputed[i] == FALSE) {
    subttl <- paste0(wdi_SE_anomalies$Country.Name[i], " - ",
                     wdi_SE_anomalies$Indicator.Name[i])
    print(plot(wdi_SE_anomalies,
               country = wdi_SE_anomalies$Country.Code[i],
               indicator = wdi_SE_anomalies$Indicator.Code[i],
               .total_threshold = 2, x.lab = subttl))
  }
}

# To plot one selected anomaly (provide country code and indicator code)
plot(wdi_SE_anomalies,
     country = "IRL", # Ireland
     indicator = "SE.PRM.ENRR", # School enrollment, primary (% gross)
     .total_threshold = 2, x.lab = "Year")
```


Running all six detection methods on this dataset with the parameters
shown in the script produced the following distribution of flags across
observations:

  -----------------------------------------------------------------------
       Methods flagging the      Number of observations
           observation           
  ------------------------------ ----------------------------------------
         6 methods (All)         0

            5 methods            11

            4 methods            2,987

            3 methods            13,131

            2 methods            29,599

             1 method            483,699

             0 method            486,483
  -----------------------------------------------------------------------

No observation was flagged unanimously by all six methods, which is
consistent with the expectation that different methods have different
sensitivities and are designed to detect different types of anomalies.
Observations flagged by a larger number of methods should be prioritized
for review, as convergence across methods provides stronger evidence of
a genuine anomaly. Anomalies flagged by only one or two methods could in
this case probably be ignored by data curators.

**Output and visualization**

The output dataset is sorted first by the number of methods that flagged
each observation, then by the absolute Z-score, so that the most
anomalous observations appear at the top. The top 20 flagged
observations are displayed as time series plots, each showing the full
series to which the flagged observation belongs. An example plot is
shown below. Plots distinguish between flagged and non-flagged
observations, and between originally observed and imputed values,
allowing reviewers to quickly assess the context of each detection.

![- Anomaly plot - Annual time
series](../images/media/image5.png){width="6.5in"
height="3.9381944444444446in"}

The full output is saved as a CSV file. This file contains every
observation in the input dataset --- flagged or not, imputed or not ---
and includes the following fields for each row: an imputation indicator,
a binary anomaly flag for each detection method, and the anomaly score
where the method produces one. This structure allows users to apply
their own prioritization criteria and to filter or rank detections in
ways suited to their specific review workflow.

### Example 2: Monthly Series of Unemployment Rate

This example applies the same workflow to a different dataset: monthly
estimates of the unemployment rate by sex and age group, sourced from
the International Labour Organisation (ILO). As of 18 April 2026, this
selection comprises 409,566 observations across 48 indicators and 68
geographies.

The primary difference from the previous example is the frequency of the
data. The time periods in the data file are provided in the format
yyyyMmm where yyyy is the year and mm is the month (01 to 12). Monthly
series have stronger and more regular seasonal patterns than annual
series, and the decomposition step must account for this. In
macroanomaly, the monthly frequency is specified via the .frequency
argument of the normalize() function, which ensures that trend and
seasonal components are removed appropriately before detection is run.
For this example, four of the six available detection methods are
applied.

```{r}
#| eval: false

#----------------------------------------------------------------------
# Anomaly detection in ILO dataset - Unemployment rate by sex and age group
#----------------------------------------------------------------------

library(macroanomaly)
library(readr)
library(collapse)
library(tidyr)
library(dplyr)

# Set default folder >>> (enter your own) <<<
setwd("C:/.../")

# Download the data from ILO website
df <- read_csv("https://rplumber.ilo.org/data/indicator/?id=UNE_DEAP_SEX_AGE_RT_M&format=.csv")

# Create a code i_code for each indicator (combine indicator, sex, age group)
df$icode <- paste0("UNE_", df$sex, "_", df$classif1)

# Drop series with insufficient number of non-missing values
min_obs <- 36
df <- df %>%
  group_by(ref_area, icode) %>%
  filter(sum(!is.na(obs_value)) >= min_obs) %>%
  ungroup()

# Normalize the dataset (detrend, impute missing values)
df |>
  macroanomaly::normalize(
    .value_col = "obs_value",
    .country_col = "ref_area",
    .indicator_col = "icode",
    .time_col = "time",
    .frequency = "monthly",
    .detrend = TRUE,
    .impute = TRUE) -> df_normalized

# Detect anomalies (4 methods)
df_normalized |>
  macroanomaly::detect(
    .method = c("tsoutlier", "capa", "zscore", "hampel"),
    .args = list(
      capa = c(.min_seg_len = 3),
      hampel = c(.trim = TRUE, .trim_sd = 1.5)),
    .additional_cols = TRUE) -> df_anomalies

df_anomalies <- df_anomalies[
  order(-df_anomalies$outlier_indicator_total,
        -df_anomalies$absZscore_zscore), ]

# Number of anomalies detected
summary(df_anomalies)
table(df_anomalies$outlier_indicator_total)

# Save output to CSV
# fn <- paste0("ILO_UNE_ANOMALIES", "_", format(Sys.Date(), "%Y-%m-%d"), ".CSV")
# write.csv(df_anomalies, file = fn, row.names = FALSE)

# Plot a sample of anomalies
for (i in 1:20) {
  if (df_anomalies$Imputed[i] == FALSE) {
    print(plot(df_anomalies,
               country = df_anomalies$ref_area[i],
               indicator = df_anomalies$icode[i],
               .total_threshold = 2, x.lab = "Month"))
  }
}
```


Running the four selected detection methods on this dataset with the
parameters shown in the script produced the following distribution of
flags across observations:

  -----------------------------------------------------------------------
       Methods flagging the      Number of observations
           observation           
  ------------------------------ ----------------------------------------
         4 methods (All)         102

            3 methods            1,512

            2 methods            2,005

             1 method            164,376

             0 method            241,571
  -----------------------------------------------------------------------

The example plot shown below shows outliers in the male unemployment
rate for population aged 35 to 44 in 2020, which may be explained by the
COVID19 pandemic.

![- Anomaly plot - Monthly time
series](../images/media/image6.png){width="6.5in"
height="4.045833333333333in"}

